GSE158422: DNA methylation in Lung squamous cell carcinoma - gene methylation enrichment analysis - sensitivity analysis

### 2023-10-04 figs/GSE158422_sensitivity.Rmd
working on sensitivity

suppressPackageStartupMessages({
  library("limma")
  library("eulerr")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library("tictoc")
  library("mitch")
  library("kableExtra")
  library("beeswarm")
})

CORES=12

@@

Start with curating the sample info.

@@

Run AL method on methylation data

```{}

# aggregation step
medf <- chragg(mval,myann,cores=CORES)

# differential analysis
dmagg <- agglimma(medf,design)
hist(dmagg$t)

head(dmagg, 30) %>%
  kbl(caption="limma gene methylation results") %>%
  kable_paper("hover", full_width = F)
# get t-stats
m <- as.data.frame(dmagg$t)
rownames(m1) <- rownames(magg1)
colnames(m1) <- "t"

# enrichment test
alres <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
hist(alres$t_mean)

@@

@@

lasig_up <- rownames(subset(lasig,t_median>0))
lasig_dn <- rownames(subset(lasig,t_median<0))

@@

v1 <- list("LA up"=lasig_up, "LA dn"=lasig_dn,
  "AL up"=alsig_up,"AL dn"=alsig_dn)

@@

tic()
dmpagg <- agg(dmp,cores=CORES)
toc()

head(dmpagg$gme_res_df,20) %>%
  kbl(caption="top gmea genes (t-test)") %>%
  kable_paper("hover", full_width = F)

@@

save.image("GSE158422_gmea.Rdata")

>>>>

suppressPackageStartupMessages({
  library("limma")
  library("eulerr")
  library("IlluminaHumanMethylation450kmanifest")
  library("IlluminaHumanMethylationEPICanno.ilm10b4.hg19")
  library("tictoc")
  library("mitch")
  library("kableExtra")
  library("beeswarm")
  library("missMethyl")
})

CORES=24

@@

Start with curating the sample info and design matrix.

@@

Run AL method on methylation data.
Record timings.

```{r,rungmea}

# aggregation step
tic()
medf <- chragg(mval,myann,cores=CORES)
toc() #51.5 sec

# differential analysis
tic()
dmagg <- agglimma(medf,design)
toc() #0.46 sec

hist(dmagg$t)

head(dmagg, 30) %>%
  kbl(caption="limma gene methylation results") %>%
  kable_paper("hover", full_width = F)

# get t-stats
m <- as.data.frame(dmagg$t)
rownames(m) <- rownames(dmagg)
colnames(m) <- "t"

# enrichment test
tic()
alres <- ttenrich(m=m,genesets=gs_symbols,cores=CORES)
toc() #1.1 sec

hist(alres$t_mean)

@@

# GSAMETH

Individual probes have a distribution that leans positive, but after aggregation to genes it is actually
negative.

```{r,rungsameth1}

# limma
runlimma <- function(mval,design,myann) {
  fit.reduced <- lmFit(mval,design)
  fit.reduced <- eBayes(fit.reduced)
  summary(decideTests(fit.reduced))
  dm <- topTable(fit.reduced,coef=4, number = Inf)
  dm <- merge(myann,dm,by=0)
  dm <- dm[order(dm$P.Value),]
  rownames(dm) <- dm$Row.names
  dm$Row.names=NULL
  return(dm)
}

dm <- runlimma(mval,design,myann)

hist(dm$t,breaks=100,xlim=c(-10,10)) ; grid()

head(dm, 10) %>% kbl() %>% kable_paper("hover", full_width = F)

sigup <- rownames(subset(dm,logFC > 0 & adj.P.Val < 0.05))
sigdn <- rownames(subset(dm,logFC < 0 & adj.P.Val < 0.05))

length(sigup)
length(sigdn)

tic()
gsaup <- gsameth(sig.cpg=sigup, all.cpg=rownames(dm), collection=gs_entrez)
gsadn <- gsameth(sig.cpg=sigdn, all.cpg=rownames(dm), collection=gs_entrez)
toc() # 12.9 sec

gsaup <- gsaup[order(gsaup$P.DE),]
head(gsaup,10) %>% kbl(caption="Top hypermethylated pathways") %>% kable_paper("hover", full_width = F)

gsadn <- gsadn[order(gsadn$P.DE),]
head(gsadn,10) %>% kbl(caption="Top hypomethylated pathways") %>% kable_paper("hover", full_width = F)

gsig_up <-  rownames(subset(gsaup,FDR<0.05))
gsig_dn <-  rownames(subset(gsadn,FDR<0.05))

@@

@@

gsig_up <-  rownames(subset(gsaup,FDR<0.05))
gsig_dn <-  rownames(subset(gsadn,FDR<0.05))

v1 <- list("AL up"=alsig_up,"AL dn"=alsig_dn,
  "GSAmeth up"=gsig_up,"GSAmeth dn"=gsig_dn)

@@

#tic()
#dmpagg <- agg(dmp,cores=CORES)
#toc()

#head(dmpagg$gme_res_df,20) %>%
#  kbl(caption="top gmea genes (t-test)") %>%
#  kable_paper("hover", full_width = F)

@@

save.image("GSE158422_sensitivity.Rdata")
